Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-19 paper.tex
add lemma:mrca_move - one rnni move to decrease rank(mrca(C))

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
Ranked trees $(T_1, r_1)$ and $(T_2, r_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
We will also refer to a ranked tree simply by tree.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.

@@

\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all
\todo{I added `non-trivial' here because the clusters corresponding to leaves cannot be put on the list.
Check that the indexing of $C_i$ is still correct.
I think we need to be explicit whether we ever use trivial clusters and I suggest that we don't.}
non-trivial
clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.

@@

Within this section we will two introduce algorithms, namely $\findpath$ and $\csort$ for computing paths between two given trees in $\rnni$.
These will be used later to prove some geometric properties of the $\rnni$ graph.
Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
\todo{LC: can we use the term max dist here?}
from a given tree.

@@

To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-1}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-1$.
\todo{In that case the number of clusters should be $n-2$.}
Trivial clusters do not need to be changed on any path between two trees which is why they are not considered in the algorithm.
First, set $\hat{T} = T$
Then for each $i$, find the node that induces $C_i$ in $\hat T$ and denote this node by $\mrca_{\hat{T}}(C_i)$ (short for \emph{most recent common ancestor}).
Update $\hat{T}$ by
\todo{This presumes  that the rank is higher.
Why?}
decreasing
the rank of $\mrca_{\hat{T}}(C_i)$ by performing $\rnni$ moves until it has the same rank $i$ as in $R$.
\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?
AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is
\todo{What is the `parent of a cluster' -- please check that you follow definitions you introduce.}
the parent of cluster
$C_i$ in $\hat{T}$ after these $\rnni$ moves.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
\todo{This kind of structure is not ideal (see my comment on the paper structure above).
If we talk about the algorithm here, all the things, including the pseudo-code should be here.
Possible solution: have algorithmic stuff first and geometry second.
If you do it like this, explain that we need the algorithms for the geometry section.}

@@

\subsection{Cluster Theorem}
\label{section:computation_cluster_theorem}

>>>>

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
We will also refer to a ranked tree simply by tree.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.

@@

@@

Within this section we will first two introduce algorithms, namely $\findpath$ and $\csort$, for computing paths between two given trees in $\rnni$.
These will be used later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
\todo{LC: can we use the term max dist here?}
from a given tree.

@@

\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
Before introducing the algorithm, we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all
\todo{Check that the indexing of $C_i$ is still correct.}
non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
\todo{LC: so we ever use trivial clusters? If so, we need to add a comment here}

To compute a path from tree $T$ to $R$ we consider clusters $C_1, \ldots, C_{n-2}$ of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
\todo{In that case the number of clusters should be $n-2$.}
First, set $\hat{T} = T$
Then for each $i$, find the node that induces $C_i$ in $\hat T$ and denote this node by $\mrca_{\hat{T}}(C_i)$ (short for \emph{most recent common ancestor}).
Update $\hat{T}$ by
\todo{This presumes  that the rank is higher.
Why?}
decreasing
the rank of $\mrca_{\hat{T}}(C_i)$ by performing $\rnni$ moves until it has the same rank $i$ as in $R$.
\todo{Do we need to explain why there is always an $\rnni$ move that moves the $\mrca$ down?
AG: We need to explain quite a bit here -- there  are many ways one can do those moves + comment above.}
Since we consider the clusters of $R$ such that the ranks of the corresponding internal nodes in $R$ increase, $\mrca_{\hat{T}}(C_i)$ is the node inducing cluster $C_i$ in $\hat{T}$ after these $\rnni$ moves.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.
\todo{This kind of structure is not ideal (see my comment on the paper structure above).
If we talk about the algorithm here, all the things, including the pseudo-code should be here.
Possible solution: have algorithmic stuff first and geometry second.
If you do it like this, explain that we need the algorithms for the geometry section.}

%BEGIN: From MareikesVisit

\todo{LC: Is it worth introducing intervals for $\rnni$ moves for the proof of the following lemma?}

\begin{lemma}
    Let $T$ be a tree and $C \subset \{1, \ldots, n\}$ a subset of taxa such that $C$ is not induced by any node in $T$.
    There is exactly one tree $T'$ that can be obtained from $T$ by one $\rnni$ move such that $\rank(\mrca_{T'}(C)) = \rank(\mrca_{T}(C)) -1 $.
\end{lemma}

\begin{proof}

    \todo{LC: revise this whole proof}

    Let $(C_1, \ldots, C_n)$ be the list representation of $T$ as introduced above.
    \todo{LC: Introduce this representation above!}
    Let $v:=mrca_T(C)$, let $C_{i+1}$ be the cluster induced by $v$, and let $w$ be the node inducing $C_i$.
    We will distinguish the case that there is an edge $(v,w)$ in $T$ from the case that there is no such edge.

    At first we assume that there is no edge $(v,w)$, which means $C_i \not \subset C_{i+1}$.
    In this case the two sets $C_i, C_{i+1}$ can exchange by a rank swap.
    Since $C \subset C_{i+1}$, this rank swap decreases the rank of $\mrca(C)$ by one and there is no $\nni$ move that could move $\mrca(C)$ down.

    Let us now consider the case that $(v,w)$ is an edge in $T$.
    The only move that could decrease the rank of $\mrca(C)$ is an $\nni$ move on the edge $(v,w)$.
    Let $C_k$ and $C_j$ be the clusters induced by the children of $C_i$.
    Then it is $C_k \cup C_j = C_i$.
    There are two different $\nni$ moves possible on $(v,w)$.
    $C_i$ can either be replaced by $C_j \cup (C_{i+1} \setminus C_i)$ or by $C_k \cup (C_{i+1} \setminus C_i)$.
    To prove that there is always exactly one $\rnni$ moves that decreases the rank of $\mrca(C)$, we need to show that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.

    Let $a \in C$.
    Then it is $a \in C_{i+1} = mrca(C)$ and it follows $a \in C_{i+1} \setminus C_i$ or $a \in C_i$.
    If $a \in C_{i+1} \setminus C_i$, $a$ is in the new set after the $\nni$ move.
    So now we will assume that $a \in C_i$.
    Then it is either $a \in C_j$ or $a \in C_k$.
    Without loss of generality we assume that $a \in C_j$.

    We need to prove that if $a \in C_j$, there is no $b \in C$ with $b \in C_k$.
    This makes sure that it is either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.
    For contradiction we assume that $a \in C_j$ and $C \in S_k$.
    We know that all sets that are subsets of $C$ already are at the same position in the current tree as in the destination tree $R$.
    Therefore, we can conclude from $a \in C_j$ and $b \in C_k$ that $C_j \cup C_k = C$.
    Hence, it holds $mrca(C) = C_i$.
    This is a contradiction to our assumption that $C_{i+1} = mrca(C)$ in the current tree.
    Therefore, it holds either $C \subset (C_j \cup (C_{i+1} \setminus C_i))$ or $C \subset (C_k \cup (C_{i+1} \setminus C_i))$.

    Concluding, there is always an $\rnni$ move on the edge $(v,w)$ for $mrca(C) = C_{i+1}, C \subsetneq C_{i+1}$ that decreases the rank of $mrca(C)$.

\end{proof}

@@
